D3064R0: Making OOTA Execution on Real Implementations be Vacuous

### 2024-01-07 ootavacuous.tex
January 7 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
Alan Stern\\\email{stern@rowland.harvard.edu} \and
The Indefatigible TBD
}
\date{December 15, 2023 (Pre-Tokyo)}
\maketitle{}

@@

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by C++
	implementers.
	But at the same time, there are no known instances of OOTA
	behavior in real C++ implementations.

	This paper refactors the commonly used condition for detecting
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and on hardware systems.
	These constraints prevent OOTA from occurring in correctly
	constructed C++ programs running on correctly constructed C++
	implementations in our current universe.

@@

@@

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle.
The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access.
In this case, both processes may access both \co{x} and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

Finally, line~23 specifies an \co{exists} clause, which gives a
condition to check for the final value of the specified local
variables.
The \co{0:} prefix specifies a variable local to \co{P0()} and
the \co{1:} prefix specifies a variable local to \co{P1()}.
The \co{/\\} specifies a boolean AND, and the \co{=} specifies
an equality comparison.

@@

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases'',\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
which may be found in
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.

@@

Some workers recommend avoiding OOTA by ordering prior relaxed
loads before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this requires real instructions be executed, consuming real
time and real electrical power to solve a strictly theoretical
problem.

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

Prior work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision process for what does and does not
constitute a causal cycle.
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.
Still others argue (perhaps correctly) that it is impossible to provide
a general definition of ``OOTA cycle'', but even accepting this raises
the question of whether a \emph{useful} definition can be formulated.

@@

The current authors deviate slightly from this tradition by instead
defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the
set of store-to-load links, but excluding intra-thread rfi links, which
are instead considered to be part of sdep.
Although this does shunt even more complexity onto the term
``semantic dependency'', it also enables us to bring to bear
laws of physics and properties of hardware onto correct real-world
C++ implementations.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}
Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.

@@

Worse yet, suppose that both \co{y} and \co{z} are zero.
Then there is no semantic dependency from either to \co{x} unless both
\co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} and \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.

@@

There is a semantic dependency from both \co{y} and \co{z} if and only
if these values do not cancel.
However, if the implementation knows that the value of \co{y} is always
the negative of \co{x}, there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were
instead the constant zero.

Finally, consider this:

@@

The two loads from \co{y} are fungible, and one should therefore expect
a compiler to replace this \co{if} statement with a single load from \co{y}.

@@

In this case, is the semantic dependency from \co{y} to \co{x},
from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
The answer is that this is a free choice on the part of the
implementation.

@@

Here, \co{collatz_cycle_min()}\footnote{
	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
takes its arbitrary-precision-integer argument and repeatedly either
divides it by two if it is even, or multiplies it by three and adds one
if it is odd.
Should the resulting sequence of arbitrary-precision integers reach
a cycle, the function returns the minimum element of that cycle.
Leaving aside the practical complication of a value of \co{y} that
consumes more than half of the available memory, there is reason to
believe that the above code fragment will always assign the value 1 to
\co{x}.
Except that this reason for belief is based on the Collatz conjecture,
which is unproven and might never be proven.

A similar example could be constructed from Fermat's Last Theorem
(FLT),\footnote{
\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
which was first stated in the year 1637 and finally proved more than
300~years later, in 1994.
This of course means that it was only after 1994 that code relying on
FLT was known not to have a semantic dependency.

@@

\begin{quote}
	\co{void foo(char c} \\
	\co{\{} \\
	\co{\ \ \ \ return c + 128;} \\
	\co{\}}
\end{quote}

@@

If empirical evidence continues to validate this assumption, it is
not possible to form causal loops based on infinite-speed propagation
of information.

@@

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even given finite speed of light, information could
propagate among computing elements in zero time.
This situation would also permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

@@

In contrast, the C++11 modification-order relation is atemporal,
because computing systems can and observably do determine the
modification order of a set of concurrent stores long after the fact.
This means that the store that executed earlier in global time
might appear later in the stored-to object's modification
order~\cite{McKenney20xxParallelProgramming}.
The atemporal nature of C++11 modification order is due to the hardware
store-buffer optimizations used by modern multicore systems.

@@

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative exection from ever committing.

@@

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, one could imagine that both the speculation
	and the squashing of failed speculation might span multiple
	threads.
	This might make sense for multiple hardware threads sharing
	a core.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

@@

Note also that some architectures provide conditional-move instructions,
which allows line~7 to avoid emitting a conditional branch.
However, such architectures would still treat a store of the
target of the conditional-move instructions as speculative until
such time as the conditional moves (and any loads that they depend
on) complete.

@@

@@

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted.

@@

For more discussion along these lines, please see
Section~\ref{sec:But What About Non-Volatile Atomics?}.

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe'')), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference interprocess communications
	links.}

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other c++ implementations (for example, those using interpreters or JITs)
might need to directly model any non-determinism.
This latter situation might motivate additional C++ projects to ban
non-determinism in favor of portable code.

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the other,
whether by working with that implementation's intermediate representations,
binaries produced by that implementation, or by tracing the program's
execution.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

Rather than look at how to prevent an implementation from forming
an OOTA cycle, this section instead looks at how to form one.
This will lead to the conclusion that a correct real-world C++
implementation cannot form an OOTA cycle.

@@

One way to accomplish this while still permitting optimization of code
containing non-volatile atomics is to model non-volatile behavior on that
of volatile atomics, but with these specific weakenings:

\begin{itemize}
\item	Loads that do not affect observable behavior can be omitted,
	along with any computations based on their return values.
\item	Given a pair of relaxed loads from the same object that can
	be reordered to be adjacent to each other, one of the pair can
	be omitted with uses of the omitted load instead using the value
	returned by the surviving load.
	This is termed \emph{load fusing}.
\item	Given a pair of relaxed stores to the same object that can
	be reordered to be adjacent to each other, the first of the pair
	can be omitted.
	This is termed \emph{store fusing}.
\item	If an expression can be proven to always result in a constant,
	that constant can be substituted for that expression.
	Note well that combining this with specialization optimizations
	can convert address and data dependencies to weaker control
	dependencies.
	Address and data dependencies order later loads and stores,
	and control dependencies order only later stores.\footnote{
		Yes, yes, the C++ standard does not mention address or
		data dependencies, and mandates control dependencies
		only indirectly.
		Such dependencies nevertheless still exist.}
\end{itemize}

@@

@@

This section has shown that an implementation that correctly
handles the memory model, observable behavior, and the as-if rule
cannot form OOTA cycles when running on real systems in this
universe.

@@

This paper has added temporal reasoning to OOTA analysis of the C++
memory model.
This work shows that correct real-world C++ implementations are inherently
incapable of producing OOTA cycles.

Therefore, if your C++ implementation produces OOTA cycles, you have bugs.
And you need to fix those bugs, regardless of whether they are in your
C++ implementation, your hardware, or your alternate universe.

@@

\begin{enumerate}
\item	Section~\ref{sec:Semantic Dependencies Are Per Execution}
	is responsible for the phrase ``on any execution'' as well
	as the proviso regarding fungible loads.
\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
	challenges are met by looking at the full context within
	a given thread.
\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
	non-determinism challenges are avoided by avoiding litmus tests
	exhibiting non-deterministic behavior.
	Again, note that most production C++ code uses compiler-based C++
	implementations, which make their non-deterministic choices at
	compile time.
\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
	optimization-choice challenges are met by using the C++
	implementation's choice of semantic dependency based on the
	optimizations that the implementation actually applied.
\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
	I/O challenges are met by again noting that I/O is observed
	behavior.
	And by avoiding litmus tests involving I/O, given that relatively
	few of the available tools handle I/O.
\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
	complexity challenges are met by looking at what the code
	actually does.
	For example, if an as-yet unproven mathematical conjecture
	holds, the code will act consistently with that conjecture,
	and vice versa.
\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
	implementer-/user-choice challenges are met by accounting for
	those choices.
	For example, if a user has specified \co{-funsigned-char},
	then the actual executions will be consistent with that choice.
	For more examples, please see
	Appendix~\ref{app:User Influence Over Language Semantics}.
\end{enumerate}

@@

There are a number of possible reactions to this situation:

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
\item	Implementations should not invent uses for values from
	non-volatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	non-volatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time an impolementation might be tempted to invent uses from
	values from non-volatile relaxed atomic loads, the loads should
	instead be fused into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead involved should two
	closely spaced loads from the same object return different values
	on the other.
\item	What other interesting situations might come to light?
\end{enumerate}

@@

Inserting \co{std::observable()} at line~12
as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler into assuming that the value of \co{local}
must necessarily be zero.

@@

By the same reasoning, Goldblatt's \co{misbhave} example that
also involves unreachability can also be addressed through insertion
of the proposed \co{std::observable()}.

@@

This section demonstrates the evaluation of sdep using
the C Bounded Model Checker
(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
This tool takes a smallish C program as input and constructs a logic
expression representing that program.
The logic expression's variables represent bits in the program's inputs
and state, and the result of that logic expression is \co{true} if that
program has a bug, for example, if it is possible for that C program to
trigger an assertion or execute an out-of-bounds array reference.
The tool then hands this expression to a SAT solver, and if this solver
finds a satisfying set of values, the tool uses that set to provide
the execution path to the bug.

>>>>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
Alan Stern\\\email{stern@rowland.harvard.edu} \and
The Indefatigible TBD
}
\date{January 7, 2024 (Pre-Tokyo)}
\maketitle{}

@@

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by C++
	implementers, and this paper summarizes some of the challenges.
	But at the same time, there are no known instances of OOTA
	behavior in real C++ implementations.

	This paper refactors the commonly used condition for detecting
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and on hardware systems.
	These constraints prevent OOTA cycles containing only volatile
	atomics from occurring in correctly constructed C++ programs
	running on correctly constructed C++ implementations in our
	current universe.
	Given current optimizations, these constraints also prevent
	OOTA cycles involving non-volatile atomics.

@@

\pagebreak

@@

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle.
The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access, in this case, \co{x}
and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

Finally, line~23 specifies an \co{exists} clause, which gives a
condition to check for the final value of the specified local
variables.
The \co{0:} prefix specifies a variable local to \co{P0()} and
the \co{1:} prefix specifies a variable local to \co{P1()}.
The \co{/\\} specifies a boolean AND, and the \co{=} specifies
an equality comparison.
If a variable appears in an \co{exists} clause, then the final value of
that variable constitutes observable behavior.

@@

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases'',\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
which also may be found in
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.

@@

Some workers recommend avoiding OOTA by ordering prior relaxed
loads before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this requires real instructions be executed, consuming real
time and real electrical power to solve a strictly theoretical
problem.
This might have been acceptable in the 1960s of my youth, but it is now
the year 2024.

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

The earliest work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision process for what does and does not
constitute a causal cycle.
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.
Still others argue (perhaps correctly) that it is impossible to provide
a general definition of ``OOTA cycle'', but this simply raises
the question of whether a \emph{useful} definition can be formulated.

@@

We deviate slightly from this tradition by instead defining an OOTA cycle
as a cycle in sdep $\cup$ rfe (or \co{acyclic(sdep ; rfe)} in \co{herd7}
notation), where rfe is the set of store-to-load links, but excluding
intra-thread rfi links, which are instead considered to be part of sdep.
Although this does shunt even more complexity onto the term
``semantic dependency'', it also enables us to bring to bear
laws of physics and properties of hardware onto correct real-world
C++ implementations.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}
Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.

@@

Worse yet, suppose that both \co{y} and \co{z} are zero.
Then there is no semantic dependency from either \co{y} or \co{z} to
\co{x} unless both \co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.

@@

There are semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation knows that the value of \co{y} is always
the negative of \co{z}, there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were
instead the constant zero, again, at least until the next assignment
to \co{x}.

Also, consider this:

@@

The two loads from \co{y} are fungible, and so a C++ implementation
could behave as if this \co{if} statement had instead been a single load
from \co{y}.

@@

In this case, is the semantic dependency from \co{y} to \co{x},
from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
The answer is that this is a free choice on the part of the
implementation.
Those wishing to change the value of \co{y} or \co{z} externally
(for example, using a debugger) would do well to avail themselves
of the \co{volatile} keyword.

\emph{Semantic dependencies can be a free choice on the part of the
C++ implementation.}

@@

Here, \co{collatz_cycle_min()}\footnote{
	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
takes its arbitrary-precision-integer argument and repeatedly either
divides it by two if it is even, or multiplies it by three and adds one
if it is odd.
Should the resulting sequence of arbitrary-precision integers reach
a cycle, the function returns the minimum element of that cycle.
Leaving aside the practical complication of a value of \co{y} that
consumes more than half of the available memory, there is reason to
believe that the above code fragment will always assign the value 1 to
\co{x}.
Except that this reason for belief is based on the Collatz conjecture,
which is currently unproven and might never be proven (or disproven).

A similar example could be constructed from Fermat's Last Theorem
(FLT),\footnote{
\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
which was first stated in the year 1637 and finally proved more than
300~years later, in 1994.
Therefore, it was only after 1994 that code relying on FLT was known
not to have a semantic dependency.

@@

\begin{quote}
	\co{void foo(char c)} \\
	\co{\{} \\
	\co{\ \ \ \ return c + 128;} \\
	\co{\}}
\end{quote}

@@

As long as empirical evidence continues to validate this assumption, it
is not possible to form causal loops based on infinite-speed propagation
of information.

@@

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even given finite speed of light, information could
propagate among computing elements in zero time.
This situation would again permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

@@

In contrast, the C++11 modification-order relation is atemporal,
because computing systems can and observably do determine the
modification order of a set of concurrent stores long after the fact
(see Appendix~\ref{app:Inter-Process Communications}).
This means that the store that executed earlier in global time
might appear later in the stored-to object's modification
order~\cite{McKenney20xxParallelProgramming}.
The atemporal nature of C++11 modification order is due to the hardware
store-buffer optimizations used by modern multicore systems.

@@

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative OOTA cycles from ever committing.

@@

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, and as noted earlier, one could imagine that both
	the speculation and the squashing of failed speculation might
	span multiple threads.
	This might make sense for multiple hardware threads sharing
	a core.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

@@

Note also that some architectures provide conditional-move instructions,
which allows line~9 to avoid emitting a conditional branch.
However, such architectures would still treat a store of the
target of the conditional-move instructions as speculative until
such time as the conditional moves (and any loads that they depend
on) complete.

@@

The key point is that a store having a semantic dependency on a given
set of loads must execute after the completion of the last of those
loads in global time.

@@

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted and on fusing of
non-volatile atomic loads from adjacent objects.

@@

\paragraph{Loads From Adjacent Non-Volatile Atomic Objects}

Suppose there are a pair of adjacent non-volatile atomic objects,
and that these objects taken together are machine-word aligned
and machine-word sized.
Suppose further that the source code contains a pair of atomic
loads, one from each of these objects, with the last of the pair
being a relaxed load.
Then the C++ implementation could choose to act as if the source code
had instead contained a single atomic load from the combined
pair of objects, with the memory-ordering semantics of the first load.

For more discussion of potential optimizations, please see
Section~\ref{sec:But What About Non-Volatile Atomics?}.

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe''), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference interprocess communications
	links.}

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug, which is
most likely a bug even in the absence of OOTA cycles.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other C++ implementations (for example, those using interpreters or JITs)
might need to directly model any non-determinism.
This latter situation might motivate additional C++ projects to ban
non-determinism in favor of portable code.

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the
other, whether by working with that implementation's intermediate
representations, binaries produced by that implementation, by tracing
the program's execution, or, in the case of dynamic tools, by actually
executing the program.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

Rather than look at how to prevent an implementation from forming an
OOTA cycle, this section instead looks at how to form one.
This will lead to the conclusion that a correct real-world C
implementation cannot form OOTA cycles,\footnote{
	Keep in mind that C does not have non-volatile atomic
	operations.}
and provide guidance on how to avoid OOTA cycles for non-volatile relaxed
atomics in C++ implementations.

@@

Keeping in mind that some early C++ implementations treated non-volatile
atomic operations as if they were volatile, one way to accomplish this
while still permitting optimization of code containing non-volatile
atomics is to model non-volatile behavior on that of volatile atomics,
but with some specific weakenings, for example:

\begin{itemize}
\item	Loads that do not affect observable behavior can be omitted,
	along with any computations based on their return values.
\item	Given a pair of relaxed loads from the same object that can
	be reordered to be adjacent to each other, one of the pair can
	be omitted with uses of the omitted load instead using the value
	returned by the surviving load.
	This is termed \emph{load fusing}.
\item	Given a pair of relaxed stores to the same object that can
	be reordered to be adjacent to each other, the first of the pair
	can be omitted.
	This is termed \emph{store fusing}.
\item	Suppose the source code contains a pair of non-volatile atomic
	loads from objects that are adjacent in memory.
	Suppose further that these loads can be reordered to be adjacent
	to each other, and that the last of the two is a relaxed load.
	Then this pair of loads can be replaced with a single load
	from both objects using the first load's memory order.
\item	Suppose the source code contains a pair of non-volatile atomic
	stores to objects that are adjacent in memory.
	Suppose further that these stores can be reordered to be adjacent
	to each other, and that the first of the two is a relaxed load.
	Then this pair of loads can be replaced with a single store
	to both objects using the last store's memory order.
\item	If an expression can be proven to always result in a constant,
	that constant can be substituted for that expression.
	Note well that combining this with specialization optimizations
	can convert address and data dependencies to weaker control
	dependencies.
	Address and data dependencies order later loads and stores,
	and control dependencies order only later stores.\footnote{
		Yes, yes, the C++ standard does not mention address or
		data dependencies, and mandates control dependencies
		only indirectly.
		Such dependencies nevertheless still exist.}
\end{itemize}

@@

Note also that many users would consider it to be an extremely unfriendly
act for the C++ implemenation to combine loads or stores to adjacent
objects where each object \co{is_lock_free()}, but the combination is not.

@@

This section has shown that an implementation that correctly handles
the memory model, observable behavior, and the as-if rule cannot form
OOTA cycles from volatile atomics when running on real systems in this
universe.
It has also presented a substantial list of optimizations that avoid
creation of OOTA cycles from non-volatile atomics.

@@

This paper has added temporal reasoning to OOTA analysis of the C++
memory model.
This work shows that correct real-world C++ implementations are inherently
incapable of producing OOTA cycles from volatile atomics.
It has also presented a substantial list of optimizations that avoid
OOTA cycles on non-volatile atomics.

@@

\begin{enumerate}
\item	Section~\ref{sec:Semantic Dependencies Are Per Execution}
	is responsible for the phrase ``on any execution'' as well
	as the proviso regarding fungible loads.
\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
	challenges are met by looking at the full context within
	a given thread.
\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
	non-determinism challenges are avoided by avoiding litmus tests
	exhibiting non-deterministic behavior.
	Again, note that most production C++ code uses compiler-based C++
	implementations, which freeze their non-deterministic choices
	into the binary at compile time.
\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
	optimization-choice challenges are met by using the C++
	implementation's choice of semantic dependency based on the
	optimizations that the implementation actually applied.
\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
	I/O challenges are met by again noting that I/O is observed
	behavior.
	And by avoiding litmus tests involving I/O, given that relatively
	few of the available tools handle I/O.
\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
	complexity challenges are met by looking at what the code
	actually does.
	For example, if an as-yet unproven mathematical conjecture
	holds, the code will act consistently with that conjecture,
	and vice versa.
\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
	implementer-/user-choice challenges are met by accounting for
	those choices.
	For example, if a user has specified \co{-funsigned-char},
	then the actual executions will be consistent with that choice.
	For more examples, please see
	Appendix~\ref{app:User Influence Over Language Semantics}.
\end{enumerate}

@@

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
\caption{Inventing Non-Boolean Atomic Loads}
\label{lst:Unused Inventing Non-Boolean Atomic Loads}
\end{listing}

This is not restricted to booleans, as can be seen in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
This example is more straightforward in that it does not make quite so
creative use of the multiple values loaded.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/inc.litmus @@
\caption{Only Atomic Increment}
\label{lst:Only Atomic Increment}
\end{listing}

But suppose that we drop the single-thread-analysis constraint for
the moment, and suppose that the C++ implementation is able to prove
that the only modification of a given atomic variable is to atomically
increment it, as is the case for \co{y} in
Listing~\ref{lst:Only Atomic Increment}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/inc-range.litmus @@
\caption{Only Atomic Increment, Optimized}
\label{lst:Only Atomic Increment, Optimized}
\end{listing}

Is a C++ implementation within its rights to behave as if the code
shown in
Listing~\ref{lst:Only Atomic Increment}
was instead as shown in
Listing~\ref{lst:Only Atomic Increment, Optimized}?
One argument in favor is that the value of \co{y} must have
passed through the value 42 in the case where \co{r0} and \co{r1}
bracket that value.
Arguments against might cite the added overhead of the invented load
in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
possibility that the check for the value 42 is intended for statistical
sampling.

There are a number of possible reactions to these situations:

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
\item	Implementations should not invent uses for values from
	non-volatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	non-volatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time an implementation might be tempted to invent uses from
	values from non-volatile relaxed atomic loads from the same
	object, the loads should instead be fused into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	What other interesting situations might come to light?
\end{enumerate}

@@

The \co{std::observable()} at line~12 of
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler, instead forcing it to assume that the
value of \co{local} might not be zero.

@@

By the same reasoning, Goldblatt's \co{misbehave} example that
also involves unreachability can also be addressed through insertion
of the proposed \co{std::observable()}.

@@

This section demonstrates the evaluation of sdep using
the C Bounded Model Checker
(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
This tool takes a smallish C program as input and constructs a logic
expression representing that program.
The logic expression's variables represent bits in the program's inputs
and state, and the result of that logic expression is \co{true} if that
program has a bug, for example, if it is possible for that C program to
trigger an assertion or execute an out-of-bounds array reference.
The tool then hands this expression to a SAT solver, and if this solver
finds a satisfying set of values, the tool uses those values provide
the execution path to the bug.
